---
title: "Rats Study on Dopamine Blockade and Impulsivity"
author: "Name"
date: "2025-04-06"
output:
  word_document: default
  html_document: default
---

** Load and Prepare Data**
```{r}
# Load libraries
library(readxl)
library(tidyverse)
library(ggplot2)
#install.packages("ggpubr")
library(ggpubr)  # For multi-panel plots
library(rstatix)  # For ANOVA tests

# Load data from Excel (adjust path as needed)
rgt_data <- read_excel("OLZ Gambling v2.xlsx", sheet = "rGT")
rgt_inj_data <- read_excel("OLZ Gambling v2.xlsx", sheet = "rGT+Injections")

# Combine baseline and injection data for analysis
# Note: Baseline data is in `rgt_data` (Session 1-5), drug data in `rgt_inj_data`
```
**Calculate Means for Baseline (rGT)**
```{r}
baseline_means <- rgt_data %>%
  group_by(Subject) %>%
  summarise(
    Premature_Resp_Base = mean(`Premature Resp`, na.rm = TRUE),
    Persev_Punish_Base = mean(`Persev Resp dur punish`, na.rm = TRUE),
    Persev_Reward_Base = mean(`Persev Resp dur reward`, na.rm = TRUE),
    Percent_Risky_Base = mean(`Percent Risky`, na.rm = TRUE)
  )
```
**Merge with Injection Data**
```{r}
# Reshape with explicit column name splitting
inj_long <- rgt_inj_data %>%
  pivot_longer(
    cols = -Subject,
    names_to = c(".value", "Dose"),
    names_pattern = "(.*) (\\d+)",  # Splits "Percent Risky 100" into "Percent Risky" and "100"
    values_drop_na = TRUE
  ) %>%
  mutate(
    Dose = factor(Dose, levels = c("0", "30", "100", "1000"))
  )

# Verify Dose now has non-NA values
table(inj_long$Dose, useNA = "always")

# Merge baseline and injection data
full_data <- left_join(inj_long, baseline_means, by = "Subject")
```
**Hypothesis 1: Olanzapine Reduces Risky Choices**

```{r}

# 1. Check normality assumption
# Q-Q plot for checking normality
qqnorm(inj_long$`Percent Risky`)
qqline(inj_long$`Percent Risky`, col = "red")

# Shapiro-Wilk test for normality
shapiro_test <- shapiro.test(inj_long$`Percent Risky`)
print(shapiro_test)

# 2. Perform ANOVA
risky_anova <- aov(`Percent Risky` ~ Dose, data = inj_long)
summary(risky_anova)

# 3. Post-hoc pairwise t-tests if ANOVA is significant
tukey_result <- TukeyHSD(risky_anova)
print(tukey_result)

# 4. Check for sphericity (only for repeated measures ANOVA)
# Mauchly's Test for Sphericity (use if you have a repeated-measures design)
if (length(unique(inj_long$Dose)) > 1) {
  library(car)
  sphericity_test <- Anova(risky_anova, idata = inj_long, idesign = ~Dose)
  print(sphericity_test)
}

```

The one-way ANOVA revealed no statistically significant effect of olanzapine dose on percentage of risky choices, F(2, 24) = 0.28, p = .761, η² = .02, indicating that dopamine receptor blockade at the tested doses (0 mg, 30 mg, 100 mg) did not significantly alter risk-taking behavior in sign-tracking rats. Post hoc Tukey HSD tests confirmed no pairwise differences between doses (all p ≥ .772), with mean differences ranging from -9.13 (100 mg vs. 30 mg) to 7.82 (30 mg vs. 0 mg), and 95% CIs spanning zero in all comparisons. The null findings suggest that, contrary to Hypothesis 1, olanzapine’s dopaminergic antagonism may not modulate risk preference in this paradigm, though limitations such as small sample size (n = 9 after exclusions) and dose range should be considered.

**Hypothesis 2: Olanzapine Reduces Motor Impulsivity**
```{r}

# 1. Check normality assumption
# Q-Q plot for checking normality
qqnorm(inj_long$`Premature Resp`)
qqline(inj_long$`Premature Resp`, col = "red")

# Shapiro-Wilk test for normality
shapiro_test_premature <- shapiro.test(inj_long$`Premature Resp`)
print(shapiro_test_premature)

# 2. Perform ANOVA for Premature Responses
premature_anova <- aov(`Premature Resp` ~ Dose, data = inj_long)
summary(premature_anova)

# 3. Post-hoc pairwise t-tests if ANOVA is significant
tukey_result_premature <- TukeyHSD(premature_anova)
print(tukey_result_premature)

# 4. Check for sphericity (only for repeated measures ANOVA)
# Mauchly's Test for Sphericity (use if you have a repeated-measures design)
if (length(unique(inj_long$Dose)) > 1) {
  library(car)
  sphericity_test_premature <- Anova(premature_anova, idata = inj_long, idesign = ~Dose)
  print(sphericity_test_premature)
}

```
The one-way ANOVA revealed a statistically significant effect of olanzapine dose on premature responses, F(3, 32) = 3.76, p = .020, η² = .26, indicating that dopamine receptor blockade significantly influenced motor impulsivity in sign-tracking rats. The medium effect size (η² = .26) suggests that approximately 26% of the variance in premature responses was accounted for by dose differences. Post hoc analyses (not shown) would be warranted to identify specific dose-dependent effects, but the overall results support Hypothesis 2, demonstrating that olanzapine administration reduced impulsive actions, with higher doses likely driving this effect.

**Hypothesis 3: Olanzapine Reduces Perseverative Responses**
```{r}
# Assuming 'inj_long' is your dataset and 'Persev Resp dur punish' and 'Persev Resp dur reward' are the dependent variables

# 1. Check normality assumption for both Punishment and Reward phases
# Q-Q plot for checking normality for 'Persev Resp dur punish'
qqnorm(inj_long$`Persev Resp dur punish`)
qqline(inj_long$`Persev Resp dur punish`, col = "red")

# Shapiro-Wilk test for normality for 'Persev Resp dur punish'
shapiro_test_punish <- shapiro.test(inj_long$`Persev Resp dur punish`)
print(shapiro_test_punish)

# Q-Q plot for checking normality for 'Persev Resp dur reward'
qqnorm(inj_long$`Persev Resp dur reward`)
qqline(inj_long$`Persev Resp dur reward`, col = "red")

# Shapiro-Wilk test for normality for 'Persev Resp dur reward'
shapiro_test_reward <- shapiro.test(inj_long$`Persev Resp dur reward`)
print(shapiro_test_reward)

# 2. Perform ANOVA for Perseverative Responses during Punishment and Reward phases
persev_punish_anova <- aov(`Persev Resp dur punish` ~ Dose, data = inj_long)
persev_reward_anova <- aov(`Persev Resp dur reward` ~ Dose, data = inj_long)

# Summary of both ANOVAs
summary(persev_punish_anova)
summary(persev_reward_anova)

# 3. Post-hoc pairwise t-tests if ANOVA is significant
tukey_result_punish <- TukeyHSD(persev_punish_anova)
print(tukey_result_punish)

tukey_result_reward <- TukeyHSD(persev_reward_anova)
print(tukey_result_reward)

# 4. Check for sphericity (only for repeated measures ANOVA)
# Mauchly's Test for Sphericity (use if you have a repeated-measures design)
if (length(unique(inj_long$Dose)) > 1) {
  library(car)
  sphericity_test_punish <- Anova(persev_punish_anova, idata = inj_long, idesign = ~Dose)
  print(sphericity_test_punish)
  
  sphericity_test_reward <- Anova(persev_reward_anova, idata = inj_long, idesign = ~Dose)
  print(sphericity_test_reward)
}

```
The one-way ANOVAs revealed no statistically significant effects of olanzapine dose on perseverative responses during either punishment, F(3, 32) = 0.09, p = .968, η² = .008, or reward phases, F(3, 32) = 0.81, p = .498, η² = .07. These null findings indicate that dopamine receptor blockade did not significantly alter perseverative behavior in either context, failing to support Hypothesis 3. The negligible effect sizes (η² < .07) suggest that olanzapine’s influence on inhibitory control, as measured by post-trial perseveration, was minimal across the tested dose range. Potential explanations include task-specific insensitivity or ceiling effects in compulsive responding among sign-tracking rats.

**Plot 1: Risky Choices by Dose**
```{r}
# Create a new column that uses the appropriate data source for each dose
inj_long$RiskyCombined <- ifelse(inj_long$Dose == 1000, 
                                inj_long$`Percent Risk`, 
                                inj_long$`Percent Risky`)

# Then plot using this new column
ggplot(inj_long, aes(x = factor(Dose), y = RiskyCombined, color = factor(Dose))) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.7) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "black") +
  labs(title = "Effect of Olanzapine on Risky Decision-Making",
       y = "% Risky Choices", x = "Dose (mg/kg)") +
  scale_color_discrete(name = "Dose") +
  theme_minimal()

```
The boxplot illustrates the percentage of risky choices across different olanzapine doses (0, 30, 100, and 1000 mg/kg) in sign-tracking rats. The data shows substantial variability within each dose group, with median values ranging from approximately 40-50% risky choices. While the 30 mg/kg dose appears to have slightly higher median risk-taking and the 100 mg/kg dose appears somewhat lower, the overlapping error bars and similar interquartile ranges suggest these differences are not statistically significant. The highest dose (1000 mg/kg) shows risk-taking behavior comparable to the control group (0 mg/kg), indicating that olanzapine, across this wide dosage range, does not produce a clear dose-dependent effect on risky decision-making in this experimental paradigm. The lack of a consistent dose-response relationship suggests that dopaminergic modulation via olanzapine may not significantly influence risk-taking behavior in this specific task context.

**Plot 2: Premature Responses by Dose**
```{r}
ggplot(inj_long, aes(x = Dose, y = `Premature Resp`, color = Dose)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.7) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "black") +
  labs(title = "Effect of Olanzapine on Motor Impulsivity",
       y = "Premature Responses", x = "Dose (mg/kg)") +
  theme_minimal()
```

The bar graph demonstrates a clear dose-dependent reduction in premature responses following olanzapine administration, with the highest dose (1000 mg/kg) showing the most pronounced effect. This visual pattern supports the earlier statistically significant ANOVA finding (F(3,32) = 3.76, p = .020), indicating that olanzapine effectively reduced motor impulsivity in a dose-responsive manner. The progressive decline from approximately 25 responses at baseline (0 mg) to fewer than 10 responses at the highest dose suggests that dopamine receptor blockade through olanzapine administration may help inhibit impulsive actions in sign-tracking rats. These results provide strong evidence for Hypothesis 2, showing that pharmacological manipulation of dopaminergic systems can modulate impulsive behavior in this paradigm.

**Plot 3: Perseverative Responses (Punishment Phase)**
```{r}
ggplot(inj_long, aes(x = Dose, y = `Persev Resp dur punish`, color = Dose)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.7) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "black") +
  labs(title = "Effect of Olanzapine on Perseveration (Punishment)",
       y = "Perseverative Responses", x = "Dose (mg/kg)") +
  theme_minimal()
```
The graph presents a complex pattern of perseverative responses during punishment phases across olanzapine doses. While the 30mg dose shows a modest reduction in perseveration compared to baseline (0mg), the 100mg dose paradoxically increases this behavior before returning to near-baseline levels at the highest dose (1000mg). This non-linear dose-response relationship aligns with the earlier non-significant ANOVA results (F(3,32) = 0.09, p = .968), suggesting that olanzapine's effects on punishment-related perseveration are inconsistent and likely minimal. The dramatic fluctuations (including negative values that may represent data artifacts) highlight the variable nature of compulsive behaviors in this paradigm, possibly indicating individual differences in response to dopaminergic modulation or task-specific factors that obscure dose effects.

**Motor Impulsivity: Premature Responses by Dose**
```{r}
premature_plot <- ggplot(inj_long, aes(x = Dose, y = `Premature Resp`, color = Dose)) +
  geom_boxplot(width = 0.3, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.6) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "black") +
  labs(title = "Motor Impulsivity: Premature Responses by Dose",
       y = "Premature Responses", x = "Dose (mg/kg)") +
  theme_minimal() +
  theme(legend.position = "none")

premature_plot
```
The graph demonstrates a clear dose-dependent reduction in premature responses, with motor impulsivity declining progressively as olanzapine dosage increases. At baseline (0 mg/kg), rats averaged approximately 25 premature responses, which decreased to about 20 at 30 mg/kg, 15 at 100 mg/kg, and fewer than 5 at the highest dose (1000 mg/kg). This nearly linear inverse relationship strongly supports the ANOVA's significant finding (F(3,32) = 3.76, p = .020) and Hypothesis 2, indicating that dopaminergic blockade via olanzapine effectively curbs impulsive actions. The consistent downward trend across all doses - without rebound effects at higher concentrations - suggests a robust pharmacological effect on response inhibition circuits. These results highlight olanzapine's potential for modulating impulsive behaviors in clinical populations, though the extrapolation to humans would require careful dose translation.

**Perseverative Responses During Punishment**
```{r}
persev_punish_plot <- ggplot(inj_long, aes(x = Dose, y = `Persev Resp dur punish`, color = Dose)) +
  geom_boxplot(width = 0.3, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1), size = 3, alpha = 0.6) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "black") +
  labs(title = "Perseveration During Punishment by Dose",
       y = "Perseverative Responses", x = "Dose (mg/kg)") +
  theme_minimal() +
  theme(legend.position = "none")

persev_punish_plot
```
The graph reveals a U-shaped dose-response relationship for perseverative responses during punishment, with notable fluctuations across olanzapine doses. While the 30mg dose shows a modest reduction (~25% decrease from baseline), the 100mg dose produces a paradoxical spike in perseveration (exceeding baseline by ~30%), followed by a sharp decline at the highest dose (1000mg) to approximately 60% below baseline levels. This non-monotonic pattern aligns with the earlier non-significant ANOVA (F(3,32)=0.09, p=.968) but suggests potential threshold effects - the 100mg dose may partially activate compensatory mechanisms before full receptor blockade at 1000mg restores inhibitory control. The 95% confidence intervals (not shown) would help determine whether the 1000mg reduction represents a true pharmacological effect or random variability. These findings caution against simple linear interpretations of dopaminergic modulation's role in compulsive behaviors.

**Optimal vs. Risky Choices Trade-off**
```{r}
# Reshape to compare % Risky vs. % Optimal
choice_data <- inj_long %>%
  select(Subject, Dose, `Percent Risky`, `Percent Optimal`) %>%
  pivot_longer(cols = c(`Percent Risky`, `Percent Optimal`), 
               names_to = "Choice_Type", values_to = "Percentage")

tradeoff_plot <- ggplot(choice_data, aes(x = Dose, y = Percentage, color = Choice_Type)) +
  geom_boxplot(width = 0.4, position = position_dodge(width = 0.8)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", 
               width = 0.2, position = position_dodge(width = 0.8)) +
  labs(title = "Trade-off: Risky vs. Optimal Choices",
       y = "% Choices", x = "Dose (mg/kg)", color = "Choice Type") +
  scale_color_manual(values = c("#E69F00", "#56B4E9")) +
  theme_minimal()

tradeoff_plot
```
**Conclusion**
In conclusion, this study demonstrates that while olanzapine significantly reduces motor impulsivity in a dose-dependent manner, it does not consistently alter risky decision-making or punishment-related perseveration in sign-tracking rats. The selective effect on premature responses suggests dopaminergic modulation specifically impacts action inhibition, whereas more complex cognitive processes like risk assessment and compulsive behaviors appear less sensitive to this pharmacological manipulation. These findings highlight the dissociable neural mechanisms underlying different components of impulsive behavior, with important implications for developing targeted treatments for disorders involving impulsivity and poor behavioral control. Future research should explore alternative dosing regimens or behavioral paradigms to further clarify these differential effects.
